Approximating the matrix exponential derivative of the CTMC generator matrix, with application to global SARS-CoV-2

### 2023-03-13 manuscript/main.tex
more writing

Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate Metropolis-type \citep{metropolis1953equation} proposals far away from the current Markov chain state.  On the one hand, this approach to proposal generation helps reduce autocorrelation between chain states and is particularly helpful within higher-dimensional state spaces.  On the other hand, numerical integration of Hamilton's equations requires repeated evaluation of the Hamiltonian potential energy's gradient, and these repeated floating-point operations may become computationally burdensome.   Using HMC for Bayesian inference, the potential energy takes the form $u(\ttheta)=-\log \pi(\ttheta)$, where $\pi(\cdot)$ is the posterior density function $\pi(\ttheta):=p(\ttheta|\Y)$ of parameter $\ttheta\in\mathbb{R}^D$ given observed data $\Y$.  In this context, a single iteration of the the St\"{o}rmer-Verlet (velocity Verlet) or leapfrog integrator \citep{leimkuhler2004simulating}, beginning at time $\tau$ and with $\epsilon>0$ small, takes the form
\begin{align}\label{eq:leap}
	\mom\left(\tau + \frac{\epsilon}{2}\right) &= \mom(\tau) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau))  \\  \nonumber
	\ttheta(\tau+\epsilon) &= \ttheta(\tau) + \epsilon \, \MM^{-1}\mom(\tau+\frac{\epsilon}{2}) \\ \nonumber
	\mom(\tau + \epsilon) &= \mom\left(\tau+\frac{\epsilon}{2}\right) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau+\epsilon))  \, ,
\end{align}
where $\mom \sim \mbox{N}_D(\Zero,\MM)$ is the Hamiltonian system's Gaussian-distributed momentum variable.  
Different strategies aim to speed up the leapfrog integrator's many log-posterior gradient evaluations $\nabla \log \pi (\ttheta)$ in different ways, as these numerical routines often represent the algorithm's computational bottleneck.  In model-specific contexts, \citet{holbrook2021massive,holbrook2022bayesian,holbrook2022viral} develop parallelization strategies, and \citet{ji2020gradients} develop dynamic programming techniques, to accelerate $\nabla \log \pi (\ttheta)$.

A small body of literature considers another approach by recognizing

>>>>

Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid,neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate Metropolis-type \citep{metropolis1953equation} proposals far away from the current Markov chain state.  On the one hand, this approach to proposal generation helps reduce autocorrelation between chain states and is particularly helpful within higher-dimensional state spaces.  On the other hand, numerical integration of Hamilton's equations requires repeated evaluation of the Hamiltonian potential energy's gradient, and these repeated floating-point operations may become computationally burdensome.   Using HMC for Bayesian inference, the potential energy takes the form $u(\ttheta)=-\log \pi(\ttheta)$, where $\pi(\cdot)$ is the posterior density function $\pi(\ttheta):=p(\ttheta|\Y)$ of parameter $\ttheta\in\mathbb{R}^D$ given observed data $\Y$.  In this context, a single iteration of the the St\"{o}rmer-Verlet (velocity Verlet) or leapfrog integrator \citep{leimkuhler2004simulating}, beginning at time $\tau$ and with $\epsilon>0$ small, takes the form
\begin{align}\label{eq:leap}
	\mom\left(\tau + \frac{\epsilon}{2}\right) &= \mom(\tau) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau))  \\  \nonumber
	\ttheta(\tau+\epsilon) &= \ttheta(\tau) + \epsilon \, \MM^{-1}\mom(\tau+\frac{\epsilon}{2}) \\ \nonumber
	\mom(\tau + \epsilon) &= \mom\left(\tau+\frac{\epsilon}{2}\right) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau+\epsilon))  \, ,
\end{align}
where $\mom \sim \mbox{N}_D(\Zero,\MM)$ is the Hamiltonian system's Gaussian-distributed momentum variable.  
Different strategies aim to speed up the leapfrog integrator's many log-posterior gradient evaluations $\nabla \log \pi (\ttheta)$ in different ways, as these numerical routines often represent the algorithm's computational bottleneck.  In model-specific contexts, \citet{holbrook2021massive,holbrook2022bayesian,holbrook2022viral} develop parallelization strategies, and \citet{ji2020gradients} develop dynamic programming techniques, to accelerate $\nabla \log \pi (\ttheta)$.

A small body of work considers another approach by replacing $\nabla \log \pi (\ttheta)$ with a suitable approximation $\widehat{\nabla} \log \pi (\ttheta)$ and recognizing that the analogue of \eqref{eq:leap} continues to satisfy path reversibility and volume preservation, two essential ingredients for well-specified HMC. It follows immediately that the resulting`surrogate trajectory HMC' continues to sample the correct target distribution $\pi(\cdot)$.   The majority of surrogate HMC methods first obtain a small sample of exact gradient evaluations and then use some model to interpolate: \citet{zhang2017precomputing} assume the approximate gradient follows a piecewise constant form across a grid; \citet{rasmussen2003gaussian,lan2016emulation} find approximations using Gaussian processes; and \citet{zhang2017hamiltonian,li2019neural} use neural networks.  Recently \textcolor{red}{MAGEE}
